Minimizing Time Required For Graduation

### 2010-12-11 paper/paper.tex
Added equation references, section references in section 5

\abstract{The Bachelor's Degree is designed to take four years to complete;
however, many students fail to complete it in this time frame. One 
factor contributing to this is poor planning when registering for classes.
We present an algorithm to assist students in constructing a schedule 
that works towards completing a degree in the shortest amount of time 
possible. The algorithm presented uses heurisitics to attempt to pick the best
class set each quarter, working towards a complete schedule quartery by 
quarter.}
\pagebreak

@@

\section{Problem Description} \label{psdesc} To make our examples concrete we will focus on the
classes offered within the Math department; however, our algorithm and model can
be extended to include other departments.

@@

\subsection{Classes and Quiz Sections} Another modeling difficulty that we
initially ignore was how to represent quiz sections for a class. Each time
a course is offered is may be offered as multiple classes. For instance, during
$F$ $MATH307$ is offered as 7 different sections all teaching the same material
during different time periods. Within a specific quarter schedule line number,
SLNs uniquely identify courses allowing us to differentiate between classes with
the same name. Continuing the above example, the SLNs of the 7 $MATH307$
instances may be 15432, 15433, 15434...

Our peer review team commented that we had not accounted for courses that
required concurrent quiz sections. First we note that the quiz sections for
a class never occur during the same time periods as the class itself (otherwise
students would not be able to attend the quiz section). Quiz sections also have
unique SLNs, to include a required quiz section we include the quiz section SLN
with the SLN for the associated class. Finally, courses such at $MATH125$ can
have multiple sections, each with several quiz sections of which only one need
be taken. We found the easiest way to represent this situation was to iterate
over the quiz sections for each class section and generate a puesdo class for
each pair of class section and quiz section that occured during the time periods
for both quiz and main sections. Fig. \ref{quizzes} contains one section
$MATH125$ being split across 3 quiz sections.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{quiz_sections} 
    \end{center}
    \caption{Splitting up MATH125} 
    \label{quizzes} 
\end{figure}

@@

Additionally, we know that not every course is offered every quarter. If
a course is not offered until the following year, we could potentially add an
entire year to the schedule. This is a situation we hope to avoid. Our heuristic
function must therefore consider how often and how far in the future a given
course is offered. For example, if we are deciding between two course offering
that are identical except one is offered once a year, and the other is offered
three times a year, we want to take the course only offered once a year, as we
won't have the opportunity to take it for another entire year.

\begin{figure} [ht]
    \begin{center}
        \includegraphics[scale=0.35]{more_prereq_tree}
    \end{center}
    \caption{A Prerequisite DAG}
    \label{prereq}
\end{figure}

@@

\subsubsection{Selection Procedure} At this point in the algorithm we know which
courses are available to a student and their relative values. We present two
ways of approaching quarterly class choice.  We will initially encode the class
selection as a binary integer programming problem in spirit of the methods
presented by Dinkel et al \cite{dinkel:scheduling}. and Pritsker et al
\cite{prisker:informs}. A second approach developed by Yamada et al
\cite{yamada:heuristic} and Pferschy et al \cite{pferschy:kcg} is presented for
dealing with very large class choices where binary LP would be difficult solve.

@@

Maximize: 
\begin{equation}
    \text{Value of Quarter} = \sum_{i=1}^m Value(c_i) * X_i
    \label{qtr_val}
\end{equation}
Subject To:
\begin{equation}
    \sum_{i=1}^m Weight(c_i) * X_i \leq l, 
    \label{weight_lim}
\end{equation}

\begin{equation}
    \forall t \in O,X_i+X_j \leq 1, \forall \text{ connected pairs } (X_i,X_j) \in t
    \label{time_con}
\end{equation}

\begin{equation}
   X_i + X_j \leq 1, \forall \text{ pairs } (X_i,X_j) \text{ where } X_i \text{and} X_j \text{ are offerings of
   the same course}
    \label{course_con}
\end{equation}

The linear program
can be interpreted as follows: \begin{itemize} \item The value of a quarter is
the total number of courses taken that quarter.  \item In each quarter, the
total number of credits cannot exceed the specified limit.  \item Only one
course can be taken in one particular time slot \item A course can only be taken
or not taken. In other words, decision variables are binary.  \item Each
non-zero decision variable in the optimal solution represents a class we will
take.  \end{itemize}

Each of the equations based on eqn.~\ref{time_con} ensure that we only take one class per 
time period per day. Also this linear program is ‘nice’ in that it always has 
feasible origin (don’t take any classes), and that the simplest formulation 
is already in standard form.

As noted in the reference literature this is actually a NP-hard
problem as described \cite{pferschy:kcg}. For versions of this problem in the
low hundreds of variables, however it is still possible to solve using modern LP
solving software \cite{yamada:heuristic}. In our case, just solving for the
optimal classes for a single major fits well into the scope of a few hundred (or
even just tens of) variables. For multiple majors or to consider class selection
for distribution and elective requirements, however the number of classes
balloons significantly and the KCG problem becomes much harder. For this version
of the problem we suggest the techniques used by Yamada et al
\cite{yamada:heuristic}, which we briefly outline below.

>>>>

\abstract{The Bachelor's Degree is designed to take four years to complete;
however, many students fail to complete it in this time frame. One 
factor contributing to this is poor planning when registering for classes.
We present an algorithm to assist students in constructing a schedule 
that works towards completing a degree in the shortest amount of time 
possible. The algorithm presented uses heuristics to attempt to pick the best
class set each quarter, working towards a complete schedule quarterly by 
quarter.}
\pagebreak

@@

\section{Problem Description} \label{pdesc} To make our examples concrete we will focus on the
classes offered within the Math department; however, our algorithm and model can
be extended to include other departments.

@@

\subsection{Classes and Quiz Sections} \label{classDefs} Another modeling
difficulty that we initially ignore was how to represent quiz sections for
a class. Each time a course is offered is may be offered as multiple classes.
For instance, during $F$ $MATH307$ is offered as 7 different sections all
teaching the same material during different time periods. Within a specific
quarter schedule line number, SLNs uniquely identify courses allowing us to
differentiate between classes with the same name. Continuing the above example,
the SLNs of the 7 $MATH307$ instances may be 15432, 15433, 15434...

Our peer review team commented that we had not accounted for courses that
required concurrent quiz sections. First we note that the quiz sections for
a class never occur during the same time periods as the class itself (otherwise
students would not be able to attend the quiz section). Quiz sections also have
unique SLNs, to include a required quiz section we include the quiz section SLN
with the SLN for the associated class. Finally, courses such at $MATH125$ can
have multiple sections, each with several quiz sections of which only one need
be taken. We found the easiest way to represent this situation was to iterate
over the quiz sections for each class section and generate a pseudo class for
each pair of class section and quiz section that occurred during the time periods
for both quiz and main sections. Fig. \ref{quizzes} contains one section
$MATH125$ being split across 3 quiz sections.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{quiz_sections} 
    \end{center}
    \caption{Splitting up MATH125} 
    \label{quizzes} 
\end{figure}

@@

Additionally, we know that not every course is offered every quarter. If
a course is not offered until the following year, we could potentially add an
entire year to the schedule. This is a situation we hope to avoid. Our heuristic
function must therefore consider how often and how far in the future a given
course is offered. For example, if we are deciding between two course offering
that are identical except one is offered once a year, and the other is offered
three times a year, we want to take the course only offered once a year, as we
won't have the opportunity to take it for another entire year.
\begin{figure} [ht]
    \begin{center}
        \includegraphics[scale=0.35]{more_prereq_tree}
    \end{center}
    \caption{A Prerequisite DAG}
    \label{prereq}
\end{figure}

@@

\subsubsection{Selection Procedure} At this point in the algorithm we know which
courses are available to a student and their relative values. We present two
ways of approaching quarterly class choice.  We will initially encode the class
selection as a binary integer programming problem in spirit of the methods
presented by Dinkel et al \cite{dinkel:scheduling}. and Pritsker et al
\cite{prisker:informs}. A second approach developed by Yamada et al
\cite{yamada:heuristic} and Pferschy et al \cite{pferschy:kcg} is presented for
dealing with very large class choices where a binary LP would be difficult
solve.

@@

Maximize: 
\begin{equation}
    \text{Value of Quarter} = \sum_{i=1}^m Value(c_i) * X_i
    \label{qtrval}
\end{equation}
Subject To:
\begin{equation}
    \sum_{i=1}^m Weight(c_i) * X_i \leq l, 
    \label{weightlim}
\end{equation}

\begin{equation}
    \forall t \in O,X_i+X_j \leq 1, \forall \text{ connected pairs } (X_i,X_j) \in t
    \label{timecon}
\end{equation}

\begin{equation}
   X_i + X_j \leq 1, \forall \text{ pairs } (X_i,X_j) \text{ where } X_i \text{and} X_j \text{ are offerings of
   the same course}
    \label{coursecon}
\end{equation}

The linear program can be interpreted as follows: \begin{itemize} \item
Eqn.~\ref{qtrval}: The value of a quarter is the total number of courses taken
that quarter. We want to maximize this value, to maximize the number of high
value classes we take each quarter. \item Eqn.~\ref{weightlim}: In each quarter,
the total number of credits cannot exceed the specified limit. $Weight(c_i)$
represents a short hand way of indicating the number of credits hours a class
takes. \item Eqn.~\ref{timecon}: Only one course can be taken in one particular
time slot, this is where $G_{conflict}$ is used. Without the prebuilt conflict
graph it would be difficult to build the set of constraints based on equation
number~ref{timecon}. \item Eqn.~\ref{coursecon}: Previously in
section~\ref{classDefs} we noted that in many cases a course is offered in
multiple sections during the same quarter. A student can only take one offering
of a course, this constraint eqn must be built for every class-quiz section
combination which represent that same course.\end{itemize}

Solving this LP will yield a set $X_i$s that are 1. We represent "taking" these
classes by moving them into the set $T$. Then we advance the current quarter,
rolling the horizon back and run the algorithm again, repeating as necessary
till we have a complete schedule.

\subsection{Alternative Formulation} As noted in the reference literature this
is actually a NP-hard problem as described \cite{pferschy:kcg}.  For versions of
this problem in the low hundreds of variables, however it is still possible to
solve using modern LP solving software \cite{yamada:heuristic}. In our case,
just solving for the optimal classes for a single major fits well into the scope
of a few hundred (or even just tens of) variables. For multiple majors or to
consider class selection for distribution and elective requirements, however the
number of classes balloons significantly and the KCG problem becomes much
harder. For this version of the problem we suggest the techniques used by Yamada
et al \cite{yamada:heuristic}, which we briefly outline below.
